
```{r}
require(sp)
require(ggmap)
require(sf)
require(rgeos)
require(tidyr)
require(dplyr)
require(rgdal)
require(stringr)
require(auk)
require(lubridate)
require(reshape2)
```

```{r}
bba1 <- readRDS("rds-files\\bba1.rds")
bba2 <- readRDS("rds-files\\bba2.rds")
comp.blocks <- read.csv("List of complete blocks for Maggie.csv")
```



Observation Coverage data
```{r}
bba1c <- bba1 %>%
  mutate(coverage = ifelse(Minimum.coverage.1 == "Minimum", 
                           1, 0))

bba1c <- bba1c %>%
  mutate(coverage = ifelse(Adequate.coverage == "Adequate", 
                                2, coverage)) %>% 
  mutate(coverage = replace_na(coverage, 0))
sumc1 <- sum(bba1c$coverage)
bba1c <- bba1c %>%  
  mutate(coverage = coverage/sumc1)
bba1c$coverage
sum(bba1c$coverage)

# atlas 2
# old bba2 coverage calculation
comp.quads <- comp.blocks %>%
  mutate(Atlas.Block.Name = str_sub(Atlas.Block.Name, 1, -4))
# comp.quads
# create adequate and minimum classification for completed blocks
# adequate is any quad with more than 3 completed blocks
# Coverage is binary in atlas 2 is binary
comp.quads <- comp.quads %>%
  group_by(Atlas.Block.Name) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  mutate(quad.status = ifelse(count > 1, 2, count))
comp.quads <- comp.quads %>% distinct()
# comp.quads
bba2c <- bba2 %>%
  mutate(QuadNameCo = str_sub(QuadNameCo, 1, -4))
  # mutate(coverage = 0) # 0 = "Nope"
bba2c <- bba2c %>%
  mutate(coverage = ifelse(QuadNameCo %in% comp.quads$Atlas.Block.Name, 1, 0))
bba2c <- bba2c %>%
  mutate(coverage = ifelse(QuadNameCo %in% comp.quads$Atlas.Block.Name
                          	[comp.quads$quad.status ==2], 2,
                          	coverage))
sumc2 <- sum(bba2c$coverage)
bba2c <- bba2c %>%
  mutate(coverage = coverage/sumc2)
bba2c$coverage
sum(bba2c$coverage)
```

Converting to meters crs
```{r}
bba1m <- bba1c %>% st_transform(crs = 32619)
bba1m$coverage
bba1m <- bba1m %>%
  mutate(centroid = st_centroid(geometry))

coordinates <- st_coordinates(bba1m$centroid)

bba1m$median.lon <- coordinates[, 1]
bba1m$median.lat <- coordinates[, 2]

# bba1m <- bba1m %>%
#   mutate(median.lat = sapply(centroid, median_lat)) %>% # centroid = median coords
#   mutate(median.lon = sapply(centroid, median_lon))
# bba1m$median.lat

# Atlas 2
bba2m <- bba2c %>% st_transform(crs = 32619)
bba2m <- bba2m %>%
  mutate(centroid = st_centroid(geometry))
coordinates <- st_coordinates(bba2m$centroid)

bba2m$median.lon <- coordinates[, 1]
bba2m$median.lat <- coordinates[, 2]
bba2m
```

Conform bba1m and bba2m so they have the same blocks
```{r}
common.blocks <- intersect(bba1m$NAME, bba2m$QuadNameCo)
bba1m <- bba1m[bba1m$NAME %in% common.blocks, ]
bba2m <- bba2m[bba2m$QuadNameCo %in% common.blocks, ]

bba1m <- bba1m[order(bba1m$NAME), ]
bba2m <- bba2m[order(bba2m$QuadNameCo), ]
bba1m
bba2m
```


Add elevation column
(This takes a couple minutes to run)
```{r}
library(FedData)
library(stars)
library(terra)
library(sf)
library(dplyr)
library(raster)
#load atlas grid and plot the blocks and quads

# amap <- read_sf(dsn = '..\\Data\\Spatial Data\\blockupdate2019', layer = 'BBA_Blocks_20190418_OldManIsland') 
ele <- get_ned(amap, label = 'ME') #DEM
# ele
ele.sf <- stars::st_as_stars(ele)
bba2m.ele <- st_transform(bba2m, st_crs(ele.sf))
ele.r <- raster::raster(ele)
avg.ele <- vector()
med.ele <- vector()
# blocks <- vector
for(i in 1:nrow(bba2m.ele)){
  block <- bba2m.ele[i, ]
  tmp <- crop(ele.r, block)
  avg <- mean(values(tmp), na.rm = TRUE)
  med <- median(values(tmp), na.rm = TRUE)
  avg.ele <- c(avg.ele, avg)
  med.ele <- c(med.ele, med)
  # blocks <- c(blocks, block)
}
med.ele
bba2m$avg.ele <- avg.ele
bba2m$med.ele <- med.ele

bba1m$avg.ele <- avg.ele
bba2m$med.ele <- med.ele
bba1m
bba2m
```
Read in Land cover Data
And create block.covs data.frame
```{r}
nlcd <- get_nlcd(amap, label = 'ME', year = '2019', dataset = 'landcover')
# nlcd.sf <- stars::st_as_stars(nlcd)
bba2m.nlcd <- st_transform(bba2m, st_crs(nlcd.sf))
# nlcd.r <- raster::raster(nlcd)
plot(nlcd.r)
nlcd.sum <- table(values(nlcd.r)) #summarize the total cells of each land cover type

# ids <- unlist(attr(nlcd.sum, 'dimnames'))
# ids
# plot(nlcd.r)
#grab the names of the categories
nlcd.cat <- data.frame(LC = unlist(attr(nlcd.r, 'dimnames'))) 


# create a dataframe with just id numbers and land cover type
# rat <- levels(nlcd)

rat <- data.frame(rat)


block.covs <- matrix(NA, nrow = nrow(bba2m.nlcd), ncol = nrow(rat)) 


for(i in 1:nrow(bba2m.nlcd)){
  block <- bba2m.nlcd[i, ]
  tmp <- crop(nlcd, block)

  tmp.sum <- table(values(tmp))
  
  tmp.sum <- data.frame(tmp.sum)
  
  tmp.sum <- tmp.sum %>%
    mutate(Var1 = as.integer(as.character(Var1)))  
  missing_ids <- setdiff(rat$ID, tmp.sum$Var1)
  additional_rows <- data.frame(Var1 = missing_ids, Freq = 0)
  tmp.sum <- rbind(tmp.sum, additional_rows)
  tmp.sum <- tmp.sum[order(tmp.sum$Var1), ]

  block.covs[i, ] <- tmp.sum$Freq
}
block.covs <- data.frame(block.covs)
colnames(block.covs) <- rat$Class

block.covs <- block.covs %>% 
  mutate(geometry = bba2m$geometry)

head(block.covs)
```

RUN FROM HERE DOWN TO AVOID RELOADING NLCD AND NED DATASETS

Initialize function for calculating north and south boundaries
```{r}
# max is also true for calculating north / east lat/lon values
calculate_boundary <- function(col, atlas, max) {
  symb <- ensym(col)
  if (max) {
    atlas <- atlas %>%
      arrange(desc({{ symb }}))
  } else {
    atlas <- atlas %>%
      arrange({{ symb }})
  }


  atlas <- atlas %>%
    slice_head(n = 10)
  
  val <- mean(atlas[[col]], na.rm = TRUE)
  
  return(val)
}

# north_boundary <- calculate_boundary('median.lat', bba2m, TRUE)
# north_boundary
```

Calculate distance to coast for each quad centroid
```{r}
# load .shp file
me_coast <- read_sf("tl_2019_us_coastline")
me_coast <-  st_transform(me_coast, 32619)


bba1m <- bba1m %>%
  mutate(d.to.coast = st_distance(centroid, st_union(me_coast)))
bba2m <- bba2m %>%
  mutate(d.to.coast = st_distance(centroid, st_union(me_coast)))

bba1m$d.to.coast <- as.numeric(bba1m$d.to.coast)
bba2m$d.to.coast <- as.numeric(bba2m$d.to.coast)
```


Initialize distribution data.frames
This chunk takes about a minute to run
```{r}
# create atlas 1 species list to iterate through
start_col <- which(names(bba1c) == "American Black Duck")
end_col <- which(names(bba1c) == "Yellow-throated Vireo")
species.a1 <- colnames(bba1c)[start_col:end_col]
# create atlas 1 species list to iterate through
start_col <- which(names(bba2c) == "Acadian Flycatcher")
end_col <- which(names(bba2c) == "Yellow Warbler")
species.a2 <- colnames(bba2c)[start_col:end_col]
# create a function to make distribution data.frames
dist_df <- function(atlas, species) 
{
  # initialize empty vectors
  centroids <- vector()
  s.lat <- vector()
  n.lat <- vector()
  min.ele <- vector()
  max.ele <- vector()
  min.coast <- vector()
  max.coast <- vector()
  # min lon
  w.lon <- vector()
  # max lon
  e.lon <- vector()
  num_blocks <- vector()
  avg.ele <- vector()
  med.ele <- vector()
  for(bird in species) {
    # only include bird species that are at least possible
    filtered.atlas <- atlas %>%
      filter(!!sym(bird) > 0)
    if (nrow(filtered.atlas) > 0) {
      #THIS LINE IS CAUSING A BUG
      center <- st_centroid(st_union(filtered.atlas$centroid))
      s.boundary <- calculate_boundary(col = 'median.lat', atlas = filtered.atlas,
                                       max = FALSE)
      n.boundary <- calculate_boundary(col = 'median.lat', atlas = filtered.atlas,
                                         max = TRUE)
      w.boundary <- calculate_boundary(col = 'median.lon', atlas = filtered.atlas,
                                       max = FALSE)
      e.boundary <- calculate_boundary(col = 'median.lon', atlas = filtered.atlas,
                                       max = TRUE)
      minimum.ele <- calculate_boundary(col = 'avg.ele', atlas = filtered.atlas,
                                       max = FALSE)
      maximum.ele <- calculate_boundary(col = 'avg.ele', atlas = filtered.atlas,
                                    max = TRUE)
      minimum.coast <- calculate_boundary(col = 'd.to.coast', atlas =
                                            filtered.atlas, max = FALSE)
      maximum.coast <- calculate_boundary(col = 'd.to.coast', atlas =
                                          filtered.atlas, max = TRUE)
      blocks <- nrow(filtered.atlas)
      # average of average elevations of each block that this bird is in
      avg <- mean(filtered.atlas$avg.ele)
    }
    
      # median of average elevations of each block that this bird is in
    med <- median(filtered.atlas$avg.ele)
    avg.ele <- c(avg.ele, avg)
    med.ele <- c(med.ele, med)
    num_blocks <- c(num_blocks, blocks)
    centroids <- c(centroids, center)
    s.lat <- c(s.lat, s.boundary)
    n.lat <- c(n.lat, n.boundary)
    w.lon <- c(w.lon, w.boundary)
    e.lon <- c(e.lon, e.boundary)
    min.ele <- c(min.ele, minimum.ele)
    max.ele <- c(max.ele, maximum.ele)
    min.coast <- c(min.coast, minimum.coast)
    max.coast <- c(max.coast, maximum.coast)

  }
  
  dist <- data.frame(
    species = species,
    s.lat = s.lat,
    n.lat = n.lat,
    num.blocks = num_blocks,
    avg.ele = avg.ele,
    med.ele = med.ele,
    w.lon = w.lon,
    e.lon = e.lon,
    min.ele = min.ele,
    max.ele = max.ele,
    min.coast = min.coast,
    max.coast = max.coast
  )
  dist$geometry <- st_sfc(centroids, crs = 32619)
  dist <- st_as_sf(dist, sf_column_name = "geometry")
  # dist$geometry
  # names(dist)
  dist <- dist %>%
    mutate(lon = st_coordinates(geometry)[,1],
           lat = st_coordinates(geometry)[,2])
  return(dist)
}

bba1m.dist <- dist_df(bba1m, species.a1)
bba2m.dist <- dist_df(bba2m, species.a2)
bba1m.dist
bba2m.dist
```




# add taxonomic categorization
```{r}
tax <- read.csv("..\\Rosenberg et al species list.csv")
head(tax)

# merge with both distribution data sets
bba1m.dist <- merge(bba1m.dist, tax, by = "species")
bba2m.dist <- merge(bba2m.dist, tax, by = "species")
head(bba1m.dist)
head(bba2m.dist)
```


Create data.frames of only species that occur in both atlases
```{r}
# create data.frames that can be merged by cutting out a whole bunch of data
common.a1 <- bba1m.dist %>%
  filter(species %in% bba2m.dist$species)

common.a2 <- bba2m.dist %>%
  filter(species %in% bba1m.dist$species)
```


Create distance to coast column
```{r}

common.a1 <- common.a1 %>%
  mutate(d.to.coast = st_distance(geometry, st_union(me_coast)))
common.a2 <- common.a2 %>%
  mutate(d.to.coast = st_distance(geometry, st_union(me_coast)))

common.a1$d.to.coast <- as.numeric(common.a1$d.to.coast)
common.a2$d.to.coast <- as.numeric(common.a2$d.to.coast)

# calculate change, and add only to common.a2
# common.a2$change_d_coast <- (common.a1$d.to.coast - common.a2$d.to.coast)/1000


# repeat process for block centroids

bba1m <- bba1m %>% 
  mutate(d.to.coast = st_distance(centroid, st_union(me_coast)))
bba2m <- bba2m %>% 
  mutate(d.to.coast = st_distance(centroid, st_union(me_coast)))


```







Export the files
```{r}
saveRDS(object = bba1m, file = "rds-files\\bba1m.rds")
saveRDS(object = bba2m, file = "rds-files\\bba2m.rds")
saveRDS(object = bba1m.dist, file = "rds-files\\bba1m_distribution.rds")
saveRDS(object = bba2m.dist, file = "rds-files\\bba2m_distribution.rds")
saveRDS(object = common.a1, file = "rds-files\\common_species1.rds")
saveRDS(object = common.a2, file = "rds-files\\common_species2.rds")
saveRDS(object = block.covs, file = "rds-files\\block.covs.rds")
saveRDS(object = block.covs, file = "rds-files\\land_coverage_by_block.rds")

```




